Personalization and probabilities: the expression of propensities in online grocery shopping

### 2017-07-28 log.rmd
adding file to be printed

## Thu 13 Jul 2017 15:29:02 BST
- starting revisions by going through ref2; 
- philosophico-methodological preramble: TODO: put this as opening paragraph

@@

## Mon 17 Jul 2017 11:46:32 BST
- still not sure how radically to restructure this. 
- the simplest order I can think of
    - A. Introduction - includes a,b,c from above
    - B. Patel section + opening description
    - C. Shopping and lists -- Miller, and list section
    - D. Archaeology of recommendations -- from 1984 -2007
    - E. Experimenting with probabilistic conversions
    - F. Conclusion
- will see if this structure works with a rough cut and paste
- re-ordering has gone ok to the end of D. But section E is really tricky -- this is the conversion event; do I include the machine learning stuff as part of D; or use it to be begin E?
    - E: shift from apriori to logistic regression; then think of good features; experimentalising/trials; infrastructural operationalisation; but then what about substitution; complementarity? All of this section is meant to be about probabilisation? And the propensities idea;    
- TODO: write short paragraph describing my overall social theory of lists (approx 200 words)
- new title: Probabilities in the data analytic conversion of social order
- TODO: play with the structuring of Section E.

@@

- an experiment for the conditional probability argument using logistic regression;
    - simulated 500k customers in terms of Taking Part data
    - simulate 52 weeks of purchase history using Tesco API data
    - build a separate model for each customer; 
    - see how that model can recommend substitutions or complementarities; these are 'new possibilities'   
- TODO: move the personalization discussion higher up, perhaps in the introduction
- rough reconstruction is ~4.2k, but haven't brought in the experiments much

@@

>>>>

## Thu 13 Jul 2017 15:29:02 BST
- starting revisions by going through ref2; 
- philosophico-methodological preramble: DONE: put this as opening paragraph

@@

## Mon 17 Jul 2017 11:46:32 BST
- still not sure how radically to restructure this. 
- the simplest order I can think of
    - A. Introduction - includes a,b,c from above
    - B. Patel section + opening description
    - C. Shopping and lists -- Miller, and list section
    - D. Archaeology of recommendations -- from 1984 -2007
    - E. Experimenting with probabilistic conversions
    - F. Conclusion
- will see if this structure works with a rough cut and paste
- re-ordering has gone ok to the end of D. But section E is really tricky -- this is the conversion event; do I include the machine learning stuff as part of D; or use it to be begin E?
    - E: shift from apriori to logistic regression; then think of good features; experimentalising/trials; infrastructural operationalisation; but then what about substitution; complementarity? All of this section is meant to be about probabilisation? And the propensities idea;    
- TODO: write short paragraph describing my overall social theory of lists (approx 200 words)
- new title: Probabilities in the data analytic conversion of social order
- DONE: play with the structuring of Section E.

@@

- an experiment for the conditional probability argument using logistic regression;
    - simulated 500k customers in terms of Taking Part data
    - simulate 52 weeks of purchase history using Tesco API data
    - build a separate model for each customer; 
    - see how that model can recommend substitutions or complementarities; these are 'new possibilities'   
- DONE: move the personalization discussion higher up, perhaps in the introduction
- rough reconstruction is ~4.2k, but haven't brought in the experiments much

@@

## Thu 27 Jul 2017 15:57:03 BST
- late in the day to be starting this -- just wanted to get the one point in about building 'good feature.'  
- brought in the self-learning substitutes discussion and the complementarities
- DONE: decide whether the hadoop and the weekly model running, etc should go in

## Fri 28 Jul 2017 12:18:30 BST
- adding the hadoop stuff as a footnote to para on local mods.  
- moved the personalization stuff into the intro
- starting to rewrite the abstract; finished a version of it. 
- DONE: go through and fix subheadings
- TODO: go through TODOs
- TODO: print

### 2017-07-28 reconstruction.rmd
adding file to be printed

# Probabilities in the data analytic conversion of social order

# A. Introduction  includes a,b,c from above

This paper philosophically reconstructs [@Dewey_2017] an online grocery ordering system undergoing an upgrade. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem the system addresses.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and shopping lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  In doing this, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations in terms of probabilities and their worldly enactments.

@@

The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture. In recent research and opinion on data practice, personalization has been seen as the linchpin of predictive modelling. The reconstruction of the recommender system suggests a different way of seeing these transformations.   Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _physical_ reality, we might come to understand the world operative probabilities belong to as what the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] where the operationality of probabilities are heightened,  might shift.

The broad implication of the argument concerns how we think about social order or regularities amidst change or transformation. Proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming for better or worse organisations, institutions, economic processes and social life. Personalization has been a pivotal element, they argue, of this transformation.  I suggest too that a transformation or conversion event is occurring, but one which might also, even in the midst of much ongoing personalization,  open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh. Even if the argument is centred on shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- particle physics for instance -- do not.

# B. Patel section + opening description

@@

# C. Shopping  lists and  a theory of social ordering

The anthropologist  Daniel Miller has argued [@Miller_2012] that  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between ideals of health and commitments to thrift). What Miller suggests here is that shopping practices attempt to resolve difference between how people think they should live and how they actually live: 'we have to watch how shopping helps resolve these discrepancies between the normative and the actual, but we also need some ideas of where the normative comes from in the first place' [@Miller_2012, 72]. Whatever theory we have of how normative social order comes about, the key argument is that shopping inhabits a space between normative/idea and actual order. Importantly, and this will be a key consideration for recommender systems, shopping is not particularly personal or individual. Grocery shopping in particular intersects forms of social order concerning family and other forms of social grouping.

From an ethnomethodological standpoint, shopping lists provide important clues to how local social order is constituted, maintained, and repaired. Social ordering is done in shopping, and the shopping list as an inscriptive device stabilises certain forms of ordering at the expense of other possible orderings.

The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of logistical flows, infrastructural orderings, and  lively negotiations around actual and normative social orders.[^30.1]  Shopping lists, therefore  are intersectional ordering devices that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

@@

# D. Archaeology of recommendations  from 1984  to 2007

@@

## Apriori as conditional probabilities

@@

Many other slight adjustments, tweaks and local modifications would probably occur around the logistic regression model. The model's recommendations themselves have propensities that need to be adjusted to the scale of values of grocery retail.  
[FEELS LIKE A JUMP HERE:  see notes 26 July 2017]

@@

There are two ways of reading this kind of statement. One is to see Popper as simply stating one of the consequencies of his physicalist understanding of probabilities as tendencies inherent in a situation. Another is to focus on the pivotal phrase 'possibilities trying to realize themselves' by asking what 'trying to realize' might mean. I would suggest that Patel's description of DunnHumby's work on the new personal relevance model is just such a 'trying to realize' in important ways.

Several times, Patel emphasized the importance of 'constructing good features' from the data, and much of her presentation concerned different attempts to construct 'good features.' A 'feature' in the context of machine learning refers to a variable included in a predictive model. A 'good feature' is one that somehow improves the accuracy, precision, specificity or any of the panoply of measures applied to machine learning models in practice. I find it interesting that the main efforts that DunnHumby made to  construct good features were not closely associated with personhood, but with the relations -- the propensities -- between things.

Patel, for instance, described  the problem of 'basket similarity,' the problem that the recommender system might recommend something similar to items already in a customer's basket. A customer might be willing to substitute a similar item for something they have already chosen, but they are more likely to accept a recommendation for an item that complements rather than duplicates items in their list.

How could the recommender system sense the possible substitutions and complementaries between things for a given customer, especially since a person's sensibilities and susceptibilities concerning groceries will be subtly shaped by wider social groups, orderings and circumstances?  It could only do that by also modelling the relation between recommendations and items already ordered to construct a new feature from the data concerning how the substitutabilities and complementarities between product.  Taking all the baskets of items purchased, rather in the way that association rules does, a new feature, what DunnyHumby termed 'self-learning substitutes' was derived. The 'self learning substitutes' feature of the relevancy model draws on another data matrix, the product similarity matrix, itself generated from all previous recommendations that have led to conversion events (that is, 'a user clicked as the response'). A customer's recommendation list might already include `Tesco Ginger Nuts 300g.`   The 'self-learning substitution model' seeks to replace list items with similar but higher value items: `Mcvitie's Ginger Moments 400g.`

This new feature is complex. Patel mentioned,  for instance, that it took the form of a 'design matrix of 14,000 columns.' This means that 14,000 new explanatory variables were added to the logistic regression model producing the recommendations. Each recommendation would be subtly re-weighted by this complex derived feature. The complexity of the model itself introduces new instabilities, new 'possibilities trying to realize themselves' and the DunnyHumby data scientists used  '`L1 regularization` to drive coefficients down to zero', or, put more simply, to exclude variables there were not contributing much to the predictions produced by the model.

@@

HERE

@@

The fact that probabilisation largely takes the form of personalization tends to obscure the re-framing of social order taking place. This ordering  depends on  the mappings that inscriptive devices such as shopping lists  display. Only because shopping is already somewhat ordered by lists can a partial re-mapping, provisionally conducive to new events, take place. The conversion entailed in this re-mapping is not just about perosnalization. It is more ontological than that since it includes relations running multi-laterally between people and things in time. It is less ontological, however, in that it is not a sudden transformation, but an ongoing process of adjusting, scaffolding, intervening and configuring orders.

>>>>

---
title: "Personalization and probabilities: the expression of propensities in online grocery shopping"
author: Adrian Mackenzie^[Sociology, Lancaster University]
date: "`r format(Sys.time(), '%B %Y')`"
fontsize: 12pt
citation_package: biber
latex_engine: xelatex
bibliography: /home/mackenza/ref_bibs/uni.bib
csl: "sage-harvard.csl"
link-citations: yes
header-includes:
    - \usepackage{setspace}
    - \doublespacing
abstract:
    "Many accounts of big data systems assume that they target individuals. Personalization, with all the risks of discrimination and bias it entails, has been the focus of inquiry. This paper suggests that personalization occurs against a more expansive background that in principle is not reducible to individual attributes and desires, but includes many different forms of relations in flux. It describes the 'personalization' of an online grocery shopping recommender system. It reconstructs an account of how a recommender system constructs an ordered list of a small number of grocery items of personal relevance for each of the millions of online grocery shoppers in  a major UK supermarket chain. Drawing on a theory of probability proposed by the philsopher of science Karl Popper, it suggests that personal relevance is inevitably bound up with relations to others, and the relations between things. Using a mixture of discourse analysis and code-based reconstruction of key elements of the recommender system, it suggests that personalization is one provisional stabilization of a much more open-ended weave of propensities associated with people and things in contemporary big data configurations.   The paper explores how in the context of recommender systems the constitutive incompleteness of shopping lists, their propensity to expand or change, becomes more important than their capacity to be personalized."
---

```{r setup2, cache=FALSE, include=FALSE}
library(knitr)
library(tint)
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
output <- opts_knit$get("rmarkdown.pandoc.to")
opts_chunk$set(warning=FALSE)
opts_chunk$set(message=FALSE)
opts_chunk$set(echo=FALSE)
opts_chunk$set(error=FALSE)
opts_chunk$set(cache=TRUE)
```

This paper philosophically reconstructs [@Dewey_2017] an online grocery ordering system undergoing an upgrade from a demographically-based recommendation system to a 'personal relevance' model. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem the system addresses.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  Throughout, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations in terms of probabilities and their worldly enactments.

@@

The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture on digital platforms associated with the web and social media. In recent research and opinion on data practice, personalization has been seen as the linchpin of predictive modelling.  Big data discourse in its promissory mode  attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. Similarly, Joseph Turow and co-authors conclude in their critical account of the transformation of retail space: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public' [@Turow_2015, 476]. Analysis of internet filter bubbles and the growth of predictive platforms in industry and government make similar points [@Pariser_2011; @ONeil_2016].

The reconstruction of the recommender system suggests a different way of seeing predictive personalization by expanding the key underpinning notion of probability.  Probabilities in practice -- in this case, in recommender system -- diverge from probability theory and indeed from statistics as a science. Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _technical_ reality, we might come to understand operative probabilities along the lines of the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] where the operationality of probabilities are heightened,  might shift focus away from personalization.

The broad implication of this expansion  concerns how we approach social order or regularities amidst change or transformation. Proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming for better or worse organisations, institutions, economic processes and social life. Personalization has been the pivotal element, they argue, of this transformation.  I suggest too that a transformation or conversion event is occurring, but one which might also, even in the midst of much ongoing personalization,  open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh. Even if the argument is centred on grocery shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- social media for instance -- do not.

# B. The conversion from demographic to personalized recommendation

@@

# C. Shopping  and  a theory of list orderings

The anthropologist  Daniel Miller has argued [@Miller_2012] that  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between the ideals of health emblematised by organic products and commitments to thrift embodied in generic products). More provocatively, Miller claims that 'shopping is largely a technology for the expression of love' (85). What Miller suggests here is that shopping practices attempt to resolve difference between how people think they should live and how they actually live: 'we have to watch how shopping helps resolve these discrepancies between the normative and the actual, but we also need some ideas of where the normative comes from in the first place' [@Miller_2012, 72]. Whatever theory we have of how normative social order comes about, the key argument is that shopping inhabits a space between normative/idea and actual order. Importantly, and this will be a key consideration for recommender systems, shopping is not particularly personal or individual. Grocery shopping in particular is woven through with forms of social order concerning family and other forms of social grouping, and their associated relations (love, etc.).

Shopping lists provide important clues to how local social order is constituted, maintained, and repaired in shopping. If social ordering is done in shopping,  then shopping list as an inscriptive device stabilises certain forms of ordering at the expense of other possible orderings. The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of logistical flows, infrastructural orderings, and  lively negotiations around actual and normative social orders.[^30.1]  Shopping lists, therefore  are intersectional ordering devices that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

@@

# D. Archaeology of recommendations:  from 1984  to 2007

@@

## `Apriori` and conditional probabilities

@@

Many other slight adjustments, tweaks and local modifications would probably occur around the logistic regression model. The model's recommendations themselves have propensities that need to be adjusted to the scale of values of grocery retail, to the available infrastructures, and to the capacities of the online grocery system to merge recommendations into a grocery order in a timely fashion.  While raw data from Tesco Online feeds into the model every hour, a recommendation list for each customer is only generated once a week. Customers only shop online every few days at the most, and in some cases, only every few weeks. To continually update the top 200 recommendations for every customer would be computationally expensive. Patel briefly mentioned  specific infrastructural elements such as `hadoop` [@ApacheSoftwareFoundation_2009]. The logistics of running a predictive model affect listing practices.  The possibility of adjusting the recommendations for every customer weekly derives from an infrastructure capable of collecting data, assimilating that data to a predictive model able to generate approximately 200 million relevant recommendations, where 'relevance' depends on a matrix of probabilities of associations between people and things that shifts in time.  `Hadoop` and its legion of 'big data' variants (`mahout`, `spark`, `hive`, `pig`, `yarn`) operationalise flows of data at an infrastructural rather than analytic scale. In many respects, the quickly glossed-over infrastructural deployment of DunnHumby's relevancy model is the primary conversion event: the logistic regression model at the heart of the recommender system is no longer an analytical device but an operational one.

@@

There are two ways of reading this statement. One is to see Popper as simply stating one of the consequencies of his physicalist understanding of probabilities as tendencies inherent in a situation. Another is to focus on the pivotal phrase 'possibilities trying to realize themselves' by asking what 'trying to realize' might mean. I would suggest that Patel's description of DunnHumby's work on the new personal relevance model is just such a 'trying to realize' in important ways.

Several times, Patel emphasized the importance of 'constructing good features' from the data, and much of her presentation concerned different attempts to construct 'good features.' A 'feature' in the context of machine learning refers to a variable included in a predictive model. A 'good feature' is one that somehow improves the accuracy, precision, specificity or any of the panoply of measures applied to machine learning models in practice. The precise meaning of 'good', however, remains open to many different attempts, experiments and variations. I find it interesting that the main efforts that DunnHumby made to  construct good features were not closely associated with personhood, but with the relations -- the propensities -- between things.

Patel, for instance, addressed the problem of 'basket similarity,' the problem that the recommender system might recommend something similar to items already in a customer's basket. A customer might be willing to substitute a similar item for something they have already chosen, but they are more likely to accept a recommendation for an item that complements rather than duplicates items in their list. How could the recommender system sense the possible substitutions and complementaries between things for a given customer, especially since a person's sensibilities and susceptibilities concerning groceries will be subtly shaped by wider social groups, orderings and circumstances?  It could only do that if had some sense of the relation between items already in the list and the field of possible recommendations. The 'personal relevance' model, therefore, sought to include the relation between recommendations and items already ordered. The data scientists constructed a new feature from the data concerning how the substitutabilities and complementarities between product.  Taking all the baskets of items purchased, rather in the way that association rules does, a new feature, what DunnyHumby termed 'self-learning substitutes' was derived. The 'self learning substitutes' feature of the relevancy model draws on another data matrix, the product similarity matrix, itself generated from all previous recommendations that have led to conversion events (that is, 'a user clicked as the response'). A customer's recommendation list might already include `Tesco Ginger Nuts 300g.`   The 'self-learning substitution model' seeks to replace list items with similar but higher value items: `Mcvitie's Ginger Moments 400g.`

This new feature is complex. Patel mentioned that it took the form of a 'design matrix of 14,000 columns.' This means that 14,000 new explanatory variables were added to the logistic regression model producing the recommendations. Each recommendation would be subtly re-weighted by this complex derived feature. The complexity of the model itself introduces new instabilities, new 'possibilities trying to realize themselves' and the DunnyHumby data scientists used  '`L1 regularization` to drive coefficients down to zero', or, put more simply, to exclude variables there were not contributing much to the predictions produced by the model.

@@

@@

The fact that probabilisation largely takes the form of personalization tends to obscure the re-framing of social order taking place. This ordering  depends on  the mappings that inscriptive devices such as shopping lists  display. Only because shopping is already somewhat ordered by lists can a partial re-mapping, provisionally conducive to new events, take place. The conversion entailed in this re-mapping is not just about personalization. It is more ontological than that since it includes relations running multi-laterally between people and things in time. It is less ontological, however, in that it is not a sudden transformation, but an ongoing process of adjusting, scaffolding, intervening and configuring orders.
